---
title: "Company Financial Modeling"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
library(caret)
```

## Data import

Place the original whitespace-delimited company dataset at `data/companies.txt`.

```{r}
set.seed(1821)

data_path <- file.path("data", "companies.txt")
if (!file.exists(data_path)) {
  stop("Missing data/companies.txt. See data/README.md for the expected file.")
}

companies <- read.table(
  data_path,
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = FALSE
) %>%
  mutate(across(where(is.character), factor))

required_columns <- c(
  "Company_Name", "Sector", "Sales", "Market_Value",
  "Assets", "Profits", "Employees"
)
missing_columns <- setdiff(required_columns, names(companies))
if (length(missing_columns) > 0) {
  stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
}

str(companies)
summary(companies)
```

## Missingness and exploratory structure

```{r}
missing_summary <- sort(colSums(is.na(companies)), decreasing = TRUE)
missing_summary
```

```{r}
ggpairs(
  companies %>%
    select(Sales, Market_Value, Assets, Profits, Employees)
)
```

## Stable transformations

The original coursework used `sign(x) * log(abs(x))` for profits, which is undefined at zero. The audited workflow uses a signed `log1p` transform instead.

```{r}
signed_log1p <- function(x) sign(x) * log1p(abs(x))

model_data <- companies %>%
  transmute(
    Sector,
    log_sales = log1p(pmax(Sales, 0)),
    log_market_value = log1p(pmax(Market_Value, 0)),
    log_assets = log1p(pmax(Assets, 0)),
    log_employees = log1p(pmax(Employees, 0)),
    signed_log_profits = signed_log1p(Profits)
  ) %>%
  filter(if_all(everything(), ~ !is.na(.)))
```

## Multivariable sales regression

All regression models use the same transformed response scale; RMSE values are therefore comparable within this section. The audited source does **not** compare RMSE from raw-sales and log-sales models as though they were on the same scale.

```{r}
control_reg <- trainControl(method = "repeatedcv", number = 5, repeats = 3)

set.seed(1821)
sales_model <- train(
  log_sales ~ log_market_value + log_assets + log_employees + signed_log_profits,
  data = model_data,
  method = "lm",
  metric = "RMSE",
  trControl = control_reg
)

sales_model
summary(sales_model$finalModel)
par(mfrow = c(2, 2))
plot(sales_model$finalModel)
par(mfrow = c(1, 1))
```

## Positive market-value model

Market value is a positive continuous quantity, so a Gamma GLM with a log link is used instead of the legacy Poisson count model.

```{r}
market_data <- companies %>%
  filter(Market_Value > 0) %>%
  transmute(
    Market_Value,
    log_sales = log1p(pmax(Sales, 0)),
    log_assets = log1p(pmax(Assets, 0)),
    log_employees = log1p(pmax(Employees, 0)),
    signed_log_profits = signed_log1p(Profits)
  ) %>%
  filter(if_all(everything(), ~ !is.na(.)))

set.seed(1821)
market_model <- train(
  Market_Value ~ .,
  data = market_data,
  method = "glm",
  family = Gamma(link = "log"),
  metric = "RMSE",
  trControl = control_reg
)

market_model
summary(market_model$finalModel)
```

## Profitability classification

`Profitable` is derived from `Profits`, so the original `Profits` variable is deliberately excluded from the predictors to avoid direct target leakage.

```{r}
classification_data <- companies %>%
  mutate(Profitable = factor(if_else(Profits > 0, "profitable", "not_profitable"))) %>%
  select(-Company_Name, -Profits) %>%
  filter(!is.na(Profitable))

set.seed(1821)
class_index <- createDataPartition(
  classification_data$Profitable,
  p = 0.75,
  list = FALSE
)
class_train <- classification_data[class_index, ]
class_test <- classification_data[-class_index, ]

control_class <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(1821)
logistic_model <- train(
  Profitable ~ .,
  data = class_train,
  method = "glm",
  family = binomial(),
  metric = "Accuracy",
  trControl = control_class
)

set.seed(1821)
knn_model <- train(
  Profitable ~ .,
  data = class_train,
  method = "knn",
  metric = "Accuracy",
  trControl = control_class,
  preProcess = c("medianImpute", "center", "scale"),
  tuneLength = 10
)
```

## Held-out classification evaluation

```{r}
logistic_predictions <- predict(logistic_model, newdata = class_test)
knn_predictions <- predict(knn_model, newdata = class_test)

confusionMatrix(logistic_predictions, class_test$Profitable)
confusionMatrix(knn_predictions, class_test$Profitable)
```

## Scope

This is an academic financial-modelling project. Results illustrate regression and classification techniques and should not be interpreted as investment advice or valuation guidance.

## Session information

```{r}
sessionInfo()
```
